Determining the Structure of Length-$k$ Steenrod Operations as $\A(r)$-Modules

### 2015-06-11 paper.tex
Add bibliography, and other stuff

@@

%TODO: frontmatter
\title{Determining the Structure of Length-$k$ Steenrod Operations as $\A(r)$-Modules}
\author{Benjamin Kraft}

@@

%TODO: more intro

\begin{defn}
  Following Welcher~\cite{TODO}, we let $L(k)$ be the subspace of $\A$ spanned by admissible monomials of length $k$.
\end{defn}

The vector space $L(k)$ may be considered as a (left or right) $\A$-module, by simply doing the multiplication in $\A$, and considering any terms which are not of length $k$ to be zero.  Similarly, we may consider $L(k)$ as a module over any of the subalgebras $\A(k)$.  $L(k)$ has a nice periodic structure, which we consider in more detail in the next section.

%TODO: future work

%TODO: structure of the paper

\section{Definitions and Background}

@@

Recall that the Steenrod squares satisfy the Adem relations:
\[ \Sq^i\Sq^j = \sum_{m=0}^{m/2}\binom{j-m-1}{i-2m}\Sq^{i+j-k}\Sq^k .\]
The binomial coefficient is taken modulo 2, and it is a standard result in combinatorics that it is nonzero if and only if for each bit of the binary representation of $i-2m$ that is 1, the corresponding bit of $j-m-1$ must also be 1.  We will use this fact extensively.

@@

\subsection{The Wood \boldmath$Z$ basis}
In addition to the Adem basis, the $Z$ basis described by Wood~\cite{TODO} will be important to our work.  Let
\[ X_n = \Sq^{2^{n+1}-2^n}\Sq^{2^{n+1}-2^{n-1}}\cdots\Sq^{2^{n+1}-1}, \]
and let $Z_n = X_n X_{n-1}\cdots X_0$.  Then Wood showed that $Z_n$ is the top element of $\A(n)$, and furthermore that the following is a basis for $\A(n)$.
\begin{defn}
  The Wood $Z$ basis for $\A(n)$ consists of the set of monomials in Steenrod squares produced by multiplying any subset of the Steenrod squares in $Z_n$ in order.
\end{defn}

@@

\section{The Commutation Relation}

@@

\section{Representatives for \boldmath$\A(n)\mmod\A(n-k)$}

@@

\section{Linear Independence}

@@

  %TODO: show somewhere that the shortest terms are in fact the shortest, i.e., that all terms of $M_{k, n-k}$ are of length at least $k+1$.
  Note that the terms 
\begin{equation}\Sq^{2^{n+1}(m_n+n)+1}\Sq^{2^n(m_{n-1}+n-1)+1}\cdots\Sq^{2^{n-k+2}(m_{n-k+1}+n-k+1)+1}\label{eq:b-term}\end{equation}
  are outside of $M_{k, n-k}$ since all elements of $\A(n-k)$ have degree smaller than the last factor of~\eqref{eq:b-term}.  Then since the elements in question may be written in the given form, we know that each has a unique term outside of $M_{k, n-k}$.  Those terms are distinct, so the terms are linearly independent.
\end{proof}

\section{Filtrations of \boldmath$L(k)$}

@@

Lemma~\ref{lem:li-vs} suggests that the set
\[ B_{n,k} = \{Sq^I \mid I = (2^{n+1}m_1, 2^n m_2, \ldots, 2^{n-k+2}m_k), m_1 \geq \cdots \geq m_k \} \]
may allow us to construct a filtration of $\Lkm$ with copies of $\A(n)\mmod\A(n-k)$ as the filtration quotients, through the following theorem.

@@

%TODO: acks

%TODO: bib

>>>>

\usepackage[maxnames=99,isbn=false,eprint=true,url=false]{biblatex}
\addbibresource{paper.bib}
\renewbibmacro{in:}{}

@@

\title{Determining the Structure of Length-$k$ Steenrod Operations as $\A(r)$-Modules}
\author{Benjamin Kraft}

@@

The structure of the Steenrod Algebra $\A$ and its various subalgebras has been the object of much study.  Consider in particular the length filtration of $\A$, with filtration quotients $L(k)$ spanned by sequences of Steenrod operations of length $k$.

%TODO: add more references for context
\begin{defn}
  Following Welcher~\cite{welcher}, we let $L(k)$ be the subspace of $\A$ spanned by admissible monomials of length $k$.
\end{defn}

The vector space $L(k)$ may be considered as a (left) $\A$-module, by simply doing the multiplication in $\A$, and considering any terms which are not of length $k$ to be zero.  Similarly, we may consider $L(k)$ as a module over any of the subalgebras $\A(k)$.  $L(k)$ has a nice periodic structure, which we consider in more detail in the next section.  Welcher~\cite{welcher} proved that $L(k)$ is a free module over the subalgebra $\A(k-1)$ generated by $\Sq^1, \Sq^2, \ldots, \Sq^{2^{k-1}}$.  We consider $L(k)$ as a module over some larger $\A(n)$, and wish to determine its stucture.

\begin{defn}
  Let $\A(n)\mmod\A(n-k)$ be the algebra $\A(n)\otimes_{\A(n-k)}\F_2$, where we consider $\F_2$ as an $\A(n-k)$ module by letting the action of all elements outside of $\F_2$ be zero.
\end{defn}

%TODO: pictures of small ones
In particular, $\A(n)\mmod\A(n-k)$ has as representatives all elements of $\A(n)$ which cannot be formed by multiplying an element of $\A(n)$ by a nontrivial element of $\A(n-k)$.  We conjecture that $L(k)$, when considered as an $\A(n)$-module, is built up out of $\A(n)\mmod\A(n-k)$, in the sense that it admits a filtration which, except for some aperiodic parts at the beginning, has each filtration quotient isomorphic to $\A(n)\mmod\A(n-k)$.

Unfortunately, we were unable to show this result in the general case.  However, we show a part of the result: that there exists a basis for such a filtration which is linearly independent.  It remains only to show that the basis has the right dimensionality.  This is more difficult than it is in the case Welcher considers, since we need to keep careful control over the aperiodic parts at the beginning.

In a few small cases, we showed that the desired result holds: the case of $L(1)$ as an $\A(n)$-module, and the case of $L(2)$ as an $\A(2)$-module.  We expect that with additional algebra, possibly on a computer, the same method can be applied to any fixed small case, and hope that a similar method can be extended to prove the general case.

Finally, we show and interesting commutation relation among elements of the Steenrod algebra: if we let
\[ X_n = \Sq^{2^{n+1}-2^n}\Sq^{2^{n+1}-2^{n-1}}\cdots\Sq^{2^{n+1}-1} \]
be defined as in Wood~\cite{wood}, then for any positive integers $m$ and $n$, 
\begin{equation}X_nX_{n-1}\Sq^{2^{n+1}m} = X_n\Sq^{2^{n+1}m}X_{n-1}.\label{eq:commutation-relation}\end{equation}
This relation is useful in proving our linear independence result, and we hope it can be useful in general.

In Section~\ref{sec:defns}, we define our terms, discuss relevant prior work, and prove some elementary propositions.  In Section~\ref{sec:commutation} we prove the commutation relation~\ref{eq:commutation-relation}.  In Section~\ref{sec:reps} we prove another lemma involving representatives for $\A(n)\mmod\A(n-k)$.  In Section~\ref{sec:li} we prove the key linear independence result.  Finally, in Section~\ref{sec:lk} we bring the results together to give a sketch of the desired proof method for the general case, and prove the result in some specific small cases.

\section{Definitions and Background}\label{sec:defns}

@@

Recall that the Steenrod squares satisfy the Adem relations:
\[ \Sq^i\Sq^j = \sum_{m=0}^{m/2}\binom{j-m-1}{i-2m}\Sq^{i+j-k}\Sq^k.\]
The binomial coefficient is taken modulo 2, and it is a standard result in combinatorics that it is nonzero if and only if for each bit of the binary representation of $i-2m$ that is 1, the corresponding bit of $j-m-1$ must also be 1.  We will use this fact extensively.

@@

\subsection{The Wood \boldmath$Z$ basis}
In addition to the Adem basis, the $Z$ basis described by Wood~\cite{wood} will be important to our work.  Let
\[ X_n = \Sq^{2^{n+1}-2^n}\Sq^{2^{n+1}-2^{n-1}}\cdots\Sq^{2^{n+1}-1}, \]
and let $Z_n = X_n X_{n-1}\cdots X_0$.  Then Wood showed that $Z_n$ is the top element of $\A(n)$, and furthermore that the following is a basis for $\A(n)$.
\begin{defn}
  The Wood $Z$ basis for $\A(n)$ consists of the set of monomials in Steenrod squares produced by multiplying any subset of the Steenrod squares in $Z_n$ in order.
\end{defn}

@@

\section{The Commutation Relation}\label{sec:commutation}

@@

\section{Representatives for \boldmath$\A(n)\mmod\A(n-k)$}\label{sec:reps}

@@

\section{Linear Independence}\label{sec:li}

@@

  Note that the terms 
\begin{equation}\Sq^{2^{n+1}(m_n+n)+1}\Sq^{2^n(m_{n-1}+n-1)+1}\cdots\Sq^{2^{n-k+2}(m_{n-k+1}+n-k+1)+1}\label{eq:b-term}\end{equation}
  are outside of $M_{k, n-k}$ since all elements of $\A(n-k)$ have degree smaller than the last factor of~\eqref{eq:b-term}.  Then since the elements in question may be written in the given form, we know that each has a unique term outside of $M_{k, n-k}$.  Those terms are distinct, so the terms are linearly independent.
\end{proof}

\section{Filtrations of \boldmath$L(k)$}\label{sec:lk}

@@

Lemma~\ref{lem:li-vs} suggests that the set
\[ B_{n,k} = \{Sq^I \mid I = (2^{n+1}m_1, 2^n m_2, \ldots, 2^{n-k+2}m_k), m_1 \geq \cdots \geq m_k \} \]
may allow us to construct a filtration of $\Lkm$ with copies of $\A(n)\mmod\A(n-k)$ as the filtration quotients, in a similar method to that which Welcher used to prove that $L(k)$ is free over $\A(k-1)$.

@@

\section{Acknowledgements}

This paper would not have been possible without the support of Mark Behrens, who proposed the problem and provided guidance along the way, and funding from the MIT Class of 1995 and Paul E. Gray Funds for UROP.

\printbibliography
